An examination of the accessibility implications of a pilot COVID-19 vaccination program in Hamilton, Ontario

### 2021-04-06 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.tex
v1 Final edits before submission to Transport Findings

  \begin{abstract}
  The province of Ontario in Canada announced the pilot for a new
  vaccination program, with designated pharmacies across the province now
  able to offer COVID-19 vaccines. The accessibility of this program
  raises questions about the cost of travel and the distribution of the
  cost among the population. In our examination of the City of Hamilton we
  find that selected sites do not serve well the rural and urban
  population of Hamilton, and that the associated cost of travel is
  expected to be disproportionally borne by lower income populations.
  Modest additions to the list of pilot sites in the city can
  substantially alleviate this inequity.
  \end{abstract}

@@

\hypertarget{research-questions-and-hypotheses}{%
\section{Research Questions and
Hypotheses}\label{research-questions-and-hypotheses}}

Along with the provision of health care facilities to treat severe cases
of COVID-19 (Pereira, C. K. V. S. Braga, et al., 2021), another front in
the fight against the pandemic is the rolling out of vaccination
programs. The Province of Ontario, in Canada, announced on April 1st
2021 the expansion of a pilot program to offer vaccines in pharmacies in
the City of
Hamilton\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
This program is in addition to dedicated vaccination centers for people
aged 70+. Twenty pharmacies in Hamilton were added to an earlier list of
325 locations in other cities across the province, and the program was
extended to people aged 55 years old and over.

Critics were swift to point out that the list of pharmacies approved for
Hamilton by the province were mostly located in lower density parts of
the city that are not well serviced by transit and are difficult to
reach by
foot\footnote{See inter alia: \url{https://twitter.com/RyanMcGreal/status/1378027149790224386?s=20} and \url{https://twitter.com/NrinderWard3/status/1378679195514060801?s=20}.}.
Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast
majority of the pharmacies are in suburban Hamilton. The issue is
somewhat less clear-cut when we consider that Hamilton's population
skews suburban (see Figure \ref{fig:population-map}). Given the target
demographic for the program, it is possible that suburban sites could be
convenient for mature adults and young old: the population aged 55 to 69
in Hamilton is approximately 58,710 suburban, 35,490 urban, and only
8,360 rural. Nevertheless, the selection of sites by the province raises
some important
questions\footnote{The decision-making process to select these sites appears to have been opaque, and the typically inert Major of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
As Yu et al. (2021) note, good geographical coverage is a key element
for a successful vaccination campaign; at the same time, siting
vaccinations sites in car-oriented locations may introduce inequities in
access.

@@

\caption{\label{fig:pharmacies-and-regions}Regions with the City of Hamilton; the location of pharmacies in pilot is shown (black triangles) and urban locations for scenario analysis (white circles)}\label{fig:pharmacies-map}
\end{figure}

@@

\caption{\label{fig:population-map}Distribution of population age 55+ in the City of Hamilton}\label{fig:population-map}
\end{figure}

\hypertarget{methods-and-data}{%
\section{Methods and Data}\label{methods-and-data}}

\hypertarget{data}{%
\section{Data}\label{data}}

@@

\hypertarget{open-hamilton}{%
\subsubsection{Open Hamilton}\label{open-hamilton}}

From the open data portal of the City of
Hamilton\footnote{\url{https://open.hamilton.ca/}} we obtained
boundaries for the city's various regions (the definition of urban,
suburban, and rural regions follows the classification of development
applications).

\hypertarget{statistics-canada}{%
\subsubsection{Statistics Canada}\label{statistics-canada}}

Population and income statistics at the level of Dissemination Areas
(DAs) were retrieved using the package \texttt{cancensus} (von Bergmann
et al., 2021). Dissemination Areas are the smallest publicly available
census geography in Canada. We use data from the 2016 Population Census.

\hypertarget{transportation-tomorrow-survey}{%
\subsubsection{Transportation Tomorrow
Survey}\label{transportation-tomorrow-survey}}

Data about modal split by age by place of residence were downloaded from
the Data Retrieval System of the Transportation Tomorrow Survey
(TTS)\footnote{\url{http://dmg.utoronto.ca/}}. The data are geocoded at
the level of Traffic Analysis Zones (TAZ).

\hypertarget{other}{%
\subsubsection{Other}\label{other}}

The locations of pharmacies in the pilot were obtained from public
records and geocoded. Three urban sites not in the program were also
identified and geocoded for comparison purposes. In addition, we
converted all recorded residential parcels in the City of Hamilton to
points on the road network. Each point includes information about the
number of residential units in the parcel.

We used the population aged 55 to 59 y.o. in each Dissemination Area to
calculate the average number of people per dwelling. This value was then
assigned proportionally to the number of dwellings per parcel. The
median total household income of the corresponding DAs was joined to the
parcels. In addition, we calculated the proportion of trips by mode from
the total number of trips by each of the three modes retrieved from the
TTS data. These proportions were joined to the parcel data based on
their corresponding TAZ. The package \texttt{r5r} (Pereira, Saraiva, et
al., 2021) was used to calculate the travel time from each parcel to all
pharmacies by three modes: car, transit, and walking. For routing
purposes we used a cutoff value of 180 min and a maximum walking
distance of 10,000 m.

Once we obtained travel time tables with population, proportion of trips
by mode, and income information, we calculated the expected travel time
\(ett\) from each parcel \(i\) to a pharmacy \(j\) as follows: \[
ett_i = p^c_i \min(tt^c_{ij}) + p^t_i \min(tt^t_{ij}) + p^w_i \min(tt^w_{ij})
\] \noindent where \(p^k_i\) is the proportion of trips by mode \(k\) in
the TAZ of parcel \(i\), and \(tt^k_{ij}\) is the vector of travel times
from parcel \(i\) to the pharmacies. In other words, the expected travel
time is the weighted sum of travel times to the nearest pharmacy, with
the weights given by the expected modal split in the TAZ.

The expected travel time \(i\) was multiplied by the population in
parcel \(i\) to obtain a measure of person-hours of travel (\(PHT\)) as
follows: \[
PHT_i = P_i\cdot ett_i
\]

@@

For comparison purposes we consider a scenario with some modest
additions to the list of pharmacies in the provincial pilot. We repeat
the analysis, but include the three urban sites shown in white circles
in Figure \ref{fig:pharmacies-and-regions}. The results of this scenario
appear in the last two columns of Table \ref{tab:distribution-results}
and the two right panels of Figure \ref{fig:results}. We begin by noting
that all income groups benefit from the addition of these three sites
with shorter mean trip durations; the most remarkable difference is the
large reduction in the disparities between residents in DAs with
different levels of income. The top-right panel of Figure
\ref{fig:results} shows that the distribution of expected travel time is
now more in line for all income groups, even if the bottom two income
quintiles still have somewhat wider spreads. With respect to the
distribution of expected travel time and person-hours of travel by
region, unsurprisingly the addition of three urban vaccination sites in
the scenario makes a difference for urban residents but not for rural
residents.

The results indicate that the locations chosen by the province for the
pilot vaccination program do not serve well urban or rural residents of
the city, and there are some important questions regarding equity of
access to the program, with a disproportionate burden in the cost of
travel falling on lower income urban populations and rural populations.
A scenario did not consider candidate sites in a systematic way.
Nonetheless, selection of three sensible urban locations does much to
alleviate disparities in the burden of transportation. On the other
hand, unlike the urban context where there are numerous candidate
locations that could be chosen for a vaccination program, there are not
many candidate locations in rural regions of the city. Increasing access
in rural Hamilton likely will involve an expansion of existing mobile
vaccination pop-up
clinics\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.

@@

\caption{\label{fig:results}Distribution of expected travel time for different population groups}\label{fig:figure-results}
\end{figure}

>>>>

  \begin{abstract}
  The Government of Ontario in Canada announced the pilot for a new
  vaccination program, with designated pharmacies across the province now
  able to offer COVID-19 vaccines. The accessibility of this program
  raises questions about the cost of travel and the distribution of the
  cost among the population. In our examination of the City of Hamilton we
  find that selected sites do not serve well rural and urban residents and
  that the associated cost of travel is expected to be disproportionally
  borne by lower income populations. Modest additions to the list of pilot
  sites in the city can substantially alleviate this inequity.
  \end{abstract}

@@

\hypertarget{questions}{%
\section{Questions}\label{questions}}

Along with the provision of health care facilities to treat severe cases
of COVID-19 (Pereira, C. K. V. S. Braga, et al., 2021), another front in
the fight against the pandemic is the rolling out of vaccination
programs. The Government of Ontario, in Canada, announced on April 1st
2021 the expansion of a pilot program to offer vaccines in pharmacies in
the City of
Hamilton\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}.
This program is in addition to dedicated vaccination centers for people
aged 70+. Twenty pharmacies in Hamilton were added to an earlier list of
325 locations in other cities across the province, and the program was
extended to people aged 55 years old and over.

Critics were swift to point out that the list of pharmacies approved for
Hamilton by the province were mostly located in lower density parts of
the city that are not well serviced by transit and are difficult to
reach by
foot\footnote{See inter alia: \url{https://twitter.com/RyanMcGreal/status/1378027149790224386?s=20} and \url{https://twitter.com/NrinderWard3/status/1378679195514060801?s=20}.}.
Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast
majority of the pharmacies are in suburban Hamilton. The issue is less
clear-cut when we consider that Hamilton's older population skews
suburban (see Figure \ref{fig:population-map}). Given the target
demographic for the program, it is possible that suburban sites could be
convenient for mature adults and the young old: the population aged 55
to 69 in Hamilton is approximately 58,710 suburban, 35,490 urban, and
only 8,360 rural. Nevertheless, the selection of sites by the province
raises some important
questions\footnote{The decision-making process to select these sites appears to have been opaque, and the Mayor of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}.
As Yu et al. (2021) note, good geographical coverage is a key element
for a successful vaccination campaign; at the same time, siting
vaccinations sites in car-oriented locations may lead to inequities in
access.

@@

\caption{\label{fig:pharmacies-and-regions}Regions with the City of Hamilton; the location of pharmacies in pilot is shown (black triangles) along with urban locations for scenario analysis (white circles).}\label{fig:pharmacies-map}
\end{figure}

@@

\caption{\label{fig:population-map}Distribution of population aged 55+ in the City of Hamilton.}\label{fig:population-map}
\end{figure}

@@

\begin{itemize}
\tightlist
\item
  Urban, suburban, and rural boundary definitions from the City of
  Hamilton\footnote{\url{https://open.hamilton.ca/}}
\item
  Population and median total household income for 2016 by Dissemination
  Area (DA) boundary using the \texttt{cancensus} package (von Bergmann
  et al., 2021)
\item
  Modal split by age for traffic analysis zones (TAZ) from the 2016
  Transportation Tomorrow Survey
  (TTS)\footnote{\url{http://dmg.utoronto.ca/}}
\item
  Locations of pilot pharmacies from public records
\item
  Locations of three additional major chain pharmacies for scenario
  testing
\item
  Residential units by land parcel
\end{itemize}

Using the population aged 55 to 59 y.o. we first calculate the average
number of people per dwelling and assign them proportionally to the
dwellings by parcel. Second, median incomes and modal splits across
walking, transit, and car are joined to the parcels. Although our data
are from 2016 (the most recent available), we assume they are largely
representative of current spatial and demographic trends in the city.
Third, we use the \texttt{r5r} (Pereira, Saraiva, et al., 2021) package
to calculate the travel time from each parcel to all pharmacies by three
modes using a cutoff value of 180 min and a maximum walking distance of
10,000 m. Once we obtained travel time tables with population,
proportion of trips by mode, and income information, we calculated the
expected travel time \(ett\) from each parcel \(i\) to a pharmacy \(j\)
as follows: \[
ett_i = p^c_i \min(tt^c_{ij}) + p^t_i \min(tt^t_{ij}) + p^w_i \min(tt^w_{ij})
\] \noindent where \(p^k_i\) is the proportion of trips by mode \(k\) in
the TAZ of parcel \(i\), and \(tt^k_{ij}\) is the vector of travel times
from parcel \(i\) to the pharmacies. The expected travel time is thus
the weighted sum of travel times to the nearest pharmacy, with the
weights given by the expected modal split in the TAZ.

The expected travel time \(i\) was multiplied by the assigned population
in parcel \(i\) to obtain a measure of person-hours of travel (\(PHT\))
as follows: \[
PHT_i = P_i\cdot ett_i
\]

@@

For comparison purposes we consider a scenario with some modest
additions to the list of pharmacies in the provincial pilot. We repeat
the analysis, but include the three urban sites shown in white circles
in Figure \ref{fig:pharmacies-and-regions}. The results of this scenario
appear in the last two columns of Table \ref{tab:distribution-results}
and the two right panels of Figure \ref{fig:results}. While all income
groups benefit from the addition of these three sites with shorter mean
trip durations, the most remarkable difference is the large reduction in
the disparities between residents in DAs with lower levels of income.
The top-right panel of Figure \ref{fig:results} shows that the
distribution of expected travel time is now more in line for all income
groups, even if the bottom two income quintiles still have somewhat
wider spreads. Unsurprisingly, the addition of three urban vaccination
sites does not have a large impact for rural residents.

The results indicate that the locations chosen by the province for the
pilot vaccination program do not serve well urban or rural residents of
the city, and there are some important questions regarding equity of
access to the program, with a disproportionate burden in the cost of
travel falling on lower income urban populations and rural populations.
Selection of three sensible urban locations does much to alleviate
disparities in the burden of transportation. On the other hand, because
there are not many candidate locations in rural parts of the city,
increasing access for residents there likely necessitates an expansion
of the existing mobile vaccination pop-up clinic
program\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.

@@

\caption{\label{fig:results}Distribution of expected travel time for different population groups.}\label{fig:figure-results}
\end{figure}
